Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-09-13 writeup/conStruct.tex
fastStr time comparison

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed.
Although \texttt{conStruct}'s computation time is invariant 
with the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it currently scales poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which is an operation of order $O(n^3)$, 
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples.
Our speed, on datasets with large sample sizes, 
could be improved by adopting rank-one updates 
to the inverse of the covariance matrix 
\citep[e.g.][]{woodbury1950,sherman_morrison1950}
when updating a sample's admixture proportions, 
which alters only a single row/column of the covariance matrix. 
However, we have not implemented this yet, 
as it would likely mean losing the ability to do efficient,  
Hamiltonian Monte Carlo sampling of our parameters.
On a desktop machine with a 4.2 GHz Intel Core i7 processor, 
an analysis of the the poplar dataset (35 samples, 22,500 loci), 
running the spatial model with 7 layers for 5,000 MCMC iterations 
took 29.8 minutes.
For the black bear dataset (84 samples, 21,000 loci) 
running the spatial model with 7 layers for 5,000 MCMC iterations, 
which was more than sufficient for convergence,
took 6.9 hours.

>>>>

\paragraph{Comparison to nonspatial model-based clustering}
Above, we showed that 
(a) the nonspatial \texttt{conStruct} model recapitulates results of 
other, commonly-used nonspatial clustering methods,
and 
(b) \texttt{conStruct} can concisely capture spatial structure, which
will be common within populations.
Given this, when should methods without spatial capability be used?
One advantage these have over \texttt{conStruct} is speed when the number of samples is large.
Although \texttt{conStruct}'s computation time is invariant 
with the number of loci included in the dataset 
(after the initial calculation of the allelic covariance), 
it currently scales poorly with number of samples.
The computationally limiting step is the inversion of the parametric covariance, 
which is an operation of order $O(n^3)$, 
whereas computation time for, e.g., STRUCTURE, 
scales linearly with number of samples.
Our speed, on datasets with large sample sizes, 
could be improved by adopting rank-one updates 
to the inverse of the covariance matrix 
\citep[e.g.][]{woodbury1950,sherman_morrison1950}
when updating a sample's admixture proportions, 
which alters only a single row/column of the covariance matrix. 
We have not implemented this yet, 
as it would likely mean losing the ability to do efficient,  
Hamiltonian Monte Carlo sampling of our parameters.

For a relatively small number of samples, 
\texttt{conStruct} can be much faster than other nonspatial clustering methods.
On a desktop machine with a 4.2 GHz Intel Core i7 processor, 
an analysis of the black bear dataset (84 samples, 21,000 loci) 
running \texttt{conStruct}'s spatial model with 4 layers for 5,000 MCMC iterations, 
which was more than sufficient for convergence,
took XXX hours.
On the same machine and dataset, 
a fastSTRUCTURE run with $K=4$ took 3.76 days.
However, for a large number of samples, 
fastSTRUCTURE (or ADMIXTURE) would be much faster.
